---
title: "Reproduced analysis of FARS data"
author: "Scott Braley"
output: 
  word_document:
   reference_docx: fars_template.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r loading_libraries}
library(tidyverse)
library(knitr)
library(ggthemes)
library(foreign)
library(dplyr)

load("../data/clean_fars.Rdata")
source("../R/fars_functions.R")
```
&nbsp;

### Results  
&nbsp;

Percentages of drivers testing positive by drug type, sex and year group
```{r table_1}
clean_fars %>%
  mutate(year_cat = cut(year, breaks = c(1999, 2002, 2006, 2010),
                        labels = c("1999-2002", "2003-2006",
                                   "2007-2010"),
                        include.lowest = TRUE, right = TRUE)) %>%
  filter(!is.na(sex)) %>%
  group_by(drug_type, sex, year_cat) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1)) %>%
  select(drug_type, sex, year_cat, perc_positive) %>%
  unite(sex_year_cat, sex, year_cat) %>%
  spread(sex_year_cat, perc_positive) %>%
  knitr::kable(col.names = c("Drug type", "F 1999-2002",
                             "F 2003-2006", "F 2007-2010",
                             "M 1999-2002", "M 2003-2006",
                             "M 2007-2010"))
```
&nbsp;

Figure 1: Prevalence of nonalcohol drugs in fatally injured drivers by year and age group
```{r figure_1, fig.height=3, fig.width=6}
temp_fars <- filter(clean_fars, drug_type != "Alcohol") %>%
  filter(!is.na(agecat)) %>%
  group_by(unique_id, year, agecat) %>%
  summarize(n_non_missing = (1/5)*sum(!is.na(positive_for_drug)),
            positive_test = 
              as.logical(sum(positive_for_drug, na.rm = TRUE))) %>%
  group_by(agecat, year) %>%
  summarize(total_obs = sum(n_non_missing),
            positive_drugs = sum(positive_test),
            perc_positive = round(100 * positive_drugs / total_obs, 1))
ggplot(temp_fars, aes(x=year, y=perc_positive, group=agecat, shape = agecat)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Positive for Nonalcohol Drugs, %") +
  theme_minimal(base_size = 10) +
  scale_shape_discrete(name="Age",
                       labels=c("<25 years", "25--44 years",
                                "45--64 years", "65 years +"))
```
&nbsp;

Figure 2: Prevalence of nonalcohol drugs in fatally injured drivers by year and drug type
```{r figure_2, fig.height=3, fig.width=6}
temp_fars <- filter(clean_fars, drug_type != "Alcohol") %>%
  filter(!is.na(drug_type)) %>%
  group_by(year, drug_type) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1)) 

ggplot(temp_fars, aes(x=year, y=perc_positive, group=drug_type, shape = drug_type)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Positive for Drugs, %") +
  theme_minimal(base_size = 10) +
  scale_shape_discrete(name="Drug Type")
```
&nbsp;

Figure 3: Prevalence of cannabinoid drugs in fatally injured drivers by year and age group
```{r figure_3, fig.height=3, fig.width=6}
temp_fars <- filter(clean_fars, drug_type == "Cannabinoids") %>%
  group_by(year, agecat) %>%
  filter(!is.na(agecat)) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1)) 

ggplot(temp_fars, aes(x=year, y=perc_positive, group=agecat, shape = agecat)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Positive for Cannabinoids, %") +
  theme_minimal(base_size = 10) +
  scale_shape_discrete(name="Age",
                       labels=c("<25 years", "25--44 years",
                                "45--64 years", "65 years +"))
```


&nbsp;

Prevalence of drugs in fally injured drivers for 1999 and 2010 by drug type
```{r table_2}
#prevalence table 1999 2010
drug_prevalence_1999 <- filter(clean_fars, year == 1999)
drug_prevalence_2010 <- filter(clean_fars, year == 2010)
drug_list <- c("Alcohol", "Narcotics", "Depressants",
               "Stimulants", "Cannabinoids", "Other")
drug_prev_99 <- lapply(drug_list, prev_test, input_data = drug_prevalence_1999)
drug_prev_99 <- dplyr::tibble(drug_prev_99) %>%
  dplyr::mutate(drug = drug_list)
drug_prev_10 <- lapply(drug_list, prev_test, input_data = drug_prevalence_2010)
drug_prev_10 <- dplyr::tibble(drug_prev_10) %>%
  dplyr::mutate(drug = drug_list)
drug_prev <- left_join(drug_prev_99, drug_prev_10, by = "drug")
drug_prev <- drug_prev[c(2,1,3)]
drug_prev  %>% knitr::kable(col.names = c("Drug type","1999","2010"))

```

&nbsp;

Statistics for testing for trend in prevalence of drugs over study years by drug type using Cochran-Armitage trend test
```{r table_3, fig.align='center'}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotics", "Depressants",
               "Stimulants", "Cannabinoids", "Other")
drug_trend_tests_ca <- lapply(drug_list, test_trend_ca)
drug_trend_tests_ca <- dplyr::bind_rows(drug_trend_tests_ca) %>%
  dplyr::mutate(drug = drug_list) %>%
  dplyr::select(drug, Z, p.value)
drug_trend_tests_ca %>% knitr::kable()
```
&nbsp;

Statistics for testing for trend in prevalence of drugs over study years by drug type using Wald test of logistic regression coefficient for year
```{r table_4, fig.align='center'}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotics", "Depressants",
               "Stimulants", "Cannabinoids", "Other")
drug_trend_tests_log_reg <- lapply(drug_list, test_trend_log_reg)
drug_trend_tests_log_reg <- dplyr::bind_rows(drug_trend_tests_log_reg) %>%
  dplyr::mutate(drug = drug_list) %>%
  dplyr::select(drug, Z, p.value)
drug_trend_tests_log_reg %>% knitr::kable()
```
&nbsp;

